---
title: "Netflix text analysis"
format: html
editor: visual
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(readr)

netflix_titles <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-04-20/netflix_titles.csv')

netflix <- netflix_titles

```

```{r warning=FALSE, message=FALSE}

library(dplyr)
library(ggplot2)
library(stringr)
library(scales)


netflix_love_war <- netflix |>
  filter(str_detect(title, "(?<!\\w)(Love|War)(?!\\w)")) |>
  select(title, type, country, release_year)

total_by_year <- netflix |>
  count(release_year, name = "total_titles")

love_war_by_year <- netflix_love_war |>
  count(release_year, name = "love_war_titles")

love_war_prop <- left_join(love_war_by_year, total_by_year, by = "release_year") |>
  mutate(prop_titles = love_war_titles / total_titles)


ggplot(love_war_prop, aes(x = release_year, y = prop_titles)) +
geom_col(fill = "pink") +
geom_line(color = "red", size = 1) +
labs(
title = "Proportion of Netflix Titles Containing 'Love' or 'War' Over Time",
x = "Release Year",
y = "Proportion of Titles"
) +
theme_minimal()
```

This plot shows the proportion of titles with the words "Love" and "War" over time. It reflects that the shows released were highly correlated with historical context. Indeed, we see that the highest peak was during the Second World War.

```{r, warning=FALSE, message=FALSE}

netflix_trends <- netflix_love_war |>
mutate(
type = factor(type),
contains_heart_battle = TRUE, 
first_word = word(title, 1)
) |>
filter(!is.na(release_year))

netflix_proportions <- netflix_trends |>
group_by(release_year, type) |>
summarise(n_titles = n(), .groups = "drop") |>
group_by(release_year) |>
mutate(prop_titles = n_titles / sum(n_titles)) |>
ungroup()


ggplot(netflix_proportions, aes(x = release_year, y = prop_titles, fill = type)) +
geom_col(position = "stack") +
scale_fill_manual(values = c("Movie" = "steelblue", "TV Show" = "orange")) +
scale_y_continuous(labels = percent_format(accuracy = 1)) +
labs(
x = "Release Year",
y = "Proportion of Titles",
fill = "Content Type",
title = "Proportion of Movies vs TV Shows on Netflix Over Time"
) +
theme_minimal()
```

This plot shows the proportion of movies and TV shows available on Netflix over time. It is interesting to see an increase in TV shows in more recent years.

```{r, warning=FALSE, message=FALSE}

netflix_genre <- netflix_titles |>
mutate(
main_genre = str_extract(listed_in, "^[^,]+"),
main_genre_upper = str_to_upper(main_genre),
is_action = str_detect(main_genre, "Action")
) |>
drop_na(release_year, main_genre_upper) |>
filter(release_year >= 1950)



top_genres <- netflix_genre |>
count(main_genre_upper, sort = TRUE) |>
slice_head(n = 6) |>
pull(main_genre_upper)



netflix_top_genres <- netflix_genre |>
filter(main_genre_upper %in% top_genres) |>
mutate(
period = case_when(
release_year < 2010 ~ "Before 2010",
release_year >= 2010 & release_year < 2020 ~ "2010–2019",
release_year >= 2020 ~ "2020–Present"
),
period = factor(period, levels = c("Before 2010", "2010–2019", "2020–Present"))
)


genre_trends <- netflix_top_genres |>
count(period, main_genre_upper, is_action) |>
group_by(period) |>
mutate(prop = n / sum(n)) |>
ungroup()

ggplot(genre_trends, aes(x = period, y = prop, fill = main_genre_upper)) +
geom_col(position = position_dodge(width = 0.8)) +
scale_y_continuous(labels = percent_format(accuracy = 1)) +
labs(
title = "Evolution of Top Netflix Genres Over Time",
subtitle = "Each genre shown separately, one color per genre",
x = "Release Period",
y = "Proportion of Titles",
fill = "Genre"
) +
theme_minimal(base_size = 9) +
theme(legend.position = "top")
```

This last plot evolution in proportion of Netflix titles per genre. Following the same theme as previously, we can see it follows the audience demand, and historical context. For example, we can see an increase from 5% to 25% of titles that are international TV shows, which emphasizes a broadening target audience.

Over time, Netflix has shifted to an increasing number of TV shows comparing to movies. This is to address increase in demand, and cultural shift towards TV shows. We also notice less of an emphasis on movies with variations of Love and War, highly associated with a specific time period (mainly, the Second World War), which makes sense when looking at the historical context. In addition, there has been a shift in genre, to reach more people. There are more international TV shows released in recent years, and fewer action movies. This shows that Netflix shows are shifting in multiple ways to address increasing public demand, and societal changes.

One thing to be mindful of, is that the number of samples is limited from the 1940s to the 1990s, which explains why the values we see are so extreme, but don't exactly reflect reality.

Shivan Bansal scrapeed this data and made it publicly available on Kaggle at the following link:

<https://www.kaggle.com/datasets/shivamb/netflix-shows>

The dataset I used lives in the Tidy Tuesday database.
